---
title: "Child Malnutrition Analysis in Kenya (DHS)"
author: "Mkulu Salim"
date: "2026-05-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Introduction

This report analyzes child malnutrition in Kenya using the Demographic and Health Survey (DHS) data.  
The focus is on three key indicators:

- Stunting (Height-for-age)
- Wasting (Weight-for-height)
- Underweight (Weight-for-age)

The analysis also examines inequalities by:
- Residence
- Wealth index
- Mother's education
- Sex of the child

---

# 2. Data Preparation.

## 2.1 Data Import.
```{r data_import}
library(haven)
library(dplyr)
library(tidyr)
library(ggplot2)

DHS <- read_dta("C:/Users/HP/Desktop/KEBR8CDT/KEBR8CFL.DTA")
```


## 2.2 Variable Selection and Weight Creation.
```{r variable_selection}
DHS_Analysis <- DHS %>%
  mutate(weight = v005 / 1000000) %>%
  select(v025, v190, v106, b4, hw70, hw71, hw72, weight)
```


## 2.3 Handling Invalid Values.
```{r invalid_values_handling}
DHS_Analysis <- DHS_Analysis %>%
  mutate(
    hw70 = ifelse(hw70 > 9000, NA, hw70),
    hw71 = ifelse(hw71 > 9000, NA, hw71),
    hw72 = ifelse(hw72 > 9000, NA, hw72)
  )
```


## 2.4 Creating Nutrition Indicators.
```{r indicators_creation}
DHS_Analysis <- DHS_Analysis %>%
  mutate(
    stunted = ifelse(is.na(hw70), NA, ifelse(hw70 < -200, 1, 0)),
    underweight = ifelse(is.na(hw71), NA, ifelse(hw71 < -200, 1, 0)),
    wasted = ifelse(is.na(hw72), NA, ifelse(hw72 < -200, 1, 0))
  )
```


## 2.5 Recoding Explanatory Variables.
```{r explanatory_variables}
DHS_Analysis <- DHS_Analysis %>%
  mutate(
    residence = ifelse(v025 == 1, "Urban", "Rural"),
    wealth_group = factor(v190,
      levels = 1:5,
      labels = c("Poorest","Poorer","Middle","Richer","Richest")
    ),
    mothers_education_level = factor(v106,
      levels = 0:3,
      labels = c("No education","Primary","Secondary","Higher")
    ),
    sex_of_child = ifelse(b4 == 1, "Male", "Female")
  )
```


# 3. Prevalence Analysis.
```{r prevalence_analysis}
prevalence_results <- DHS_Analysis %>%
  summarise(
    Stunting = weighted.mean(stunted, weight, na.rm=TRUE)*100,
    Wasting = weighted.mean(wasted, weight, na.rm=TRUE)*100,
    Underweight = weighted.mean(underweight, weight, na.rm=TRUE)*100,
    .groups = "drop"
  ) %>%
  mutate(across(everything(), ~round(.,1)))

prevalence_results
```


## Table Format.
```{r table_format}
prevalence_table <- prevalence_results %>%
  pivot_longer(
    cols = everything(),
    names_to = "Indicator",
    values_to = "Prevalence"
  )

prevalence_table
```

# 4. Inequality Analysis

## 4.1 By Residence
```{r residence_analysis}
residence_analysis <- DHS_Analysis %>%
  group_by(residence) %>%
  summarise(
    Stunting = weighted.mean(stunted, weight, na.rm=TRUE)*100,
    Wasting = weighted.mean(wasted, weight, na.rm=TRUE)*100,
    Underweight = weighted.mean(underweight, weight, na.rm=TRUE)*100,
    .groups = "drop"
  )

residence_analysis
```


## 4.2 By Wealth Index.
```{r wealthindex_analysis}
wealth_analysis <- DHS_Analysis %>%
  group_by(wealth_group) %>%
  summarise(
    Stunting = weighted.mean(stunted, weight, na.rm=TRUE)*100,
    Wasting = weighted.mean(wasted, weight, na.rm=TRUE)*100,
    Underweight = weighted.mean(underweight, weight, na.rm=TRUE)*100,
    .groups = "drop"
  )

wealth_analysis
```


## 4.3 By Mother's Education.
```{r mothers_education_analysis}
education_analysis <- DHS_Analysis %>%
  group_by(mothers_education_level) %>%
  summarise(
    Stunting = weighted.mean(stunted, weight, na.rm=TRUE)*100,
    Wasting = weighted.mean(wasted, weight, na.rm=TRUE)*100,
    Underweight = weighted.mean(underweight, weight, na.rm=TRUE)*100,
    .groups = "drop"
  )

education_analysis
```


## 4.4 By Sex of Child.
```{r sex_of_child_analysis}
sex_analysis <- DHS_Analysis %>%
  group_by(sex_of_child) %>%
  summarise(
    Stunting = weighted.mean(stunted, weight, na.rm=TRUE)*100,
    Wasting = weighted.mean(wasted, weight, na.rm=TRUE)*100,
    Underweight = weighted.mean(underweight, weight, na.rm=TRUE)*100,
    .groups = "drop"
  )

sex_analysis
```


# 5. Combined Analysis.

## 5.1 Wealth × Residence.
```{r wealth_x_residence}
combined_wr <- DHS_Analysis %>%
  group_by(residence, wealth_group) %>%
  summarise(
    Stunting = weighted.mean(stunted, weight, na.rm=TRUE)*100,
    Wasting = weighted.mean(wasted, weight, na.rm=TRUE)*100,
    Underweight = weighted.mean(underweight, weight, na.rm=TRUE)*100,
    .groups = "drop"
  ) %>%
  mutate(across(c(Stunting, Wasting, Underweight), ~round(.,1)))

combined_wr
```


#  6. Visualization.

## 6.1 Overall Prevalence.
```{r overall_prevalence}
ggplot(prevalence_table,
       aes(x = Indicator, y = Prevalence, fill = Indicator)) +
  geom_col() +
  labs(title = "Prevalence of Child Malnutrition",
       y = "Prevalence (%)", x = "") +
  theme_minimal()
```


## 6.2 Stunting by Wealth.
```{r stunting_by_wealth}
ggplot(wealth_analysis,
       aes(x = wealth_group, y = Stunting, fill = wealth_group)) +
  geom_col() +
  labs(title = "Stunting by Wealth Index",
       y = "Prevalence (%)") +
  theme_minimal()
```


## 6.3 Stunting by Residence.
```{r stunting_by_residence}
ggplot(residence_analysis,
       aes(x = residence, y = Stunting, fill = residence)) +
  geom_col() +
  labs(
    title = "Stunting by Residence",
    y = "Prevalence (%)",
    x = ""
  ) +
  theme_minimal()
```


## 6.3 Combined Inequality.
```{r combined_inequality}
ggplot(combined_wr,
       aes(x = wealth_group, y = Stunting, fill = residence)) +
  geom_col(position = "dodge") +
  labs(title = "Stunting by Wealth and Residence",
       y = "Prevalence (%)") +
  theme_minimal()
```


## Wasting
```{r wasting}
ggplot(wealth_analysis,
       aes(x = wealth_group, y = Wasting, fill = wealth_group))+
  geom_col()+
  labs(title="Wasting by Wealth", y="Prevalence (%)")+
  theme_minimal()
```


## Underweight
```{r underweight}
ggplot(wealth_analysis,
       aes(x = wealth_group, y = Underweight, fill = wealth_group))+
  geom_col()+
  labs(title="Underweight by Wealth", y="Prevalence (%)")+
  theme_minimal()
```


# 7. Discussion.
The analysis shows that stunting is the most prevalent form of malnutrition in Kenya.
Clear inequalities exist across socio-economic groups.
Children in rural areas and poorer households experience higher levels of malnutrition.
Combined analysis highlights compounded inequality, where children in poor rural households are the most affected.


# 8. Conclusion
Child malnutrition remains a major public health issue in Kenya.
Targeted interventions should focus on disadvantaged populations, especially poor and rural households.