---
title: "Child Malnutrition Analysis in Kenya (DHS)"
author: "Mkulu Salim"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# 1. Introduction

This report analyzes child malnutrition in Kenya using the Demographic and Health Survey (DHS) data.  
The focus is on three key indicators:

- Stunting (Height-for-age)
- Wasting (Weight-for-height)
- Underweight (Weight-for-age)

The analysis also examines inequalities by:
- Residence
- Wealth index
- Mother's education
- Sex of the child

---

# 2. Data Preparation

## 2.1 Data Import
```{r data_import}
library(haven)
library(dplyr)
library(tidyr)
library(ggplot2)

DHS <- read_dta("C:/Users/HP/Desktop/KEBR8CDT/KEBR8CFL.DTA")
```


## 2.2 Variable Selection and Weight Creation
```{r variable_selection}
DHS_Analysis <- DHS %>%
  mutate(weight = v005 / 1000000) %>%
  select(v025, v190, v106, b4, hw70, hw71, hw72, weight)
```


## 2.3 Handling Invalid Values
```{r invalid_values_handling}
DHS_Analysis <- DHS_Analysis %>%
  mutate(
    hw70 = ifelse(hw70 > 9000, NA, hw70),
    hw71 = ifelse(hw71 > 9000, NA, hw71),
    hw72 = ifelse(hw72 > 9000, NA, hw72)
  )
```


## 2.4 Creating Nutrition Indicators
```{r indicators_creation}
DHS_Analysis <- DHS_Analysis %>%
  mutate(
    stunted = ifelse(is.na(hw70), NA, ifelse(hw70 < -200, 1, 0)),
    underweight = ifelse(is.na(hw71), NA, ifelse(hw71 < -200, 1, 0)),
    wasted = ifelse(is.na(hw72), NA, ifelse(hw72 < -200, 1, 0))
  )
```


## 2.5 Recoding Explanatory Variables
```{r explanatory_variables}
DHS_Analysis <- DHS_Analysis %>%
  mutate(
    residence = ifelse(v025 == 1, "Urban", "Rural"),
    wealth_group = factor(v190,
      levels = 1:5,
      labels = c("Poorest","Poorer","Middle","Richer","Richest")
    ),
    mothers_education_level = factor(v106,
      levels = 0:3,
      labels = c("No education","Primary","Secondary","Higher")
    ),
    sex_of_child = ifelse(b4 == 1, "Male", "Female")
  )
```


# 3. Prevalence Analysis
```{r prevalence_analysis}
prevalence_results <- DHS_Analysis %>%
  summarise(
    Stunting = weighted.mean(stunted, weight, na.rm=TRUE)*100,
    Wasting = weighted.mean(wasted, weight, na.rm=TRUE)*100,
    Underweight = weighted.mean(underweight, weight, na.rm=TRUE)*100,
    .groups = "drop"
  ) %>%
  mutate(across(everything(), ~round(.,1)))

prevalence_results
```


## Table Format
```{r table_format}
prevalence_table <- prevalence_results %>%
  pivot_longer(
    cols = everything(),
    names_to = "Indicator",
    values_to = "Prevalence"
  )

prevalence_table
```

# 4. Inequality Analysis

## 4.1 By Residence
```{r residence_analysis}
residence_analysis <- DHS_Analysis %>%
  group_by(residence) %>%
  summarise(
    Stunting = weighted.mean(stunted, weight, na.rm=TRUE)*100,
    Wasting = weighted.mean(wasted, weight, na.rm=TRUE)*100,
    Underweight = weighted.mean(underweight, weight, na.rm=TRUE)*100,
    .groups = "drop"
  )

residence_analysis
```


## 4.2 By Wealth Index
```{r wealthindex_analysis}
wealth_analysis <- DHS_Analysis %>%
  group_by(wealth_group) %>%
  summarise(
    Stunting = weighted.mean(stunted, weight, na.rm=TRUE)*100,
    Wasting = weighted.mean(wasted, weight, na.rm=TRUE)*100,
    Underweight = weighted.mean(underweight, weight, na.rm=TRUE)*100,
    .groups = "drop"
  )

wealth_analysis
```


## 4.3 By Mother's Education
```{r mothers_education_analysis}
education_analysis <- DHS_Analysis %>%
  group_by(mothers_education_level) %>%
  summarise(
    Stunting = weighted.mean(stunted, weight, na.rm=TRUE)*100,
    Wasting = weighted.mean(wasted, weight, na.rm=TRUE)*100,
    Underweight = weighted.mean(underweight, weight, na.rm=TRUE)*100,
    .groups = "drop"
  )

education_analysis
```


## 4.4 By Sex of Child
```{r sex_of_child_analysis}
sex_analysis <- DHS_Analysis %>%
  group_by(sex_of_child) %>%
  summarise(
    Stunting = weighted.mean(stunted, weight, na.rm=TRUE)*100,
    Wasting = weighted.mean(wasted, weight, na.rm=TRUE)*100,
    Underweight = weighted.mean(underweight, weight, na.rm=TRUE)*100,
    .groups = "drop"
  )

sex_analysis
```


# 5. Combined Analysis

## 5.1 Wealth × Residence
```{r wealth_x_residence}
combined_wr <- DHS_Analysis %>%
  group_by(residence, wealth_group) %>%
  summarise(
    Stunting = weighted.mean(stunted, weight, na.rm=TRUE)*100,
    Wasting = weighted.mean(wasted, weight, na.rm=TRUE)*100,
    Underweight = weighted.mean(underweight, weight, na.rm=TRUE)*100,
    .groups = "drop"
  ) %>%
  mutate(across(c(Stunting, Wasting, Underweight), ~round(.,1)))

combined_wr
```


#  6. Visualization

## 6.1 Overall Prevalence
```{r overall_prevalence}
ggplot(prevalence_table,
       aes(x = Indicator, y = Prevalence, fill = Indicator)) +
  geom_col() +
  labs(title = "Prevalence of Child Malnutrition",
       y = "Prevalence (%)", x = "") +
  theme_minimal()
```


## 6.2 Stunting by Wealth
```{r stunting_by_wealth}
ggplot(wealth_analysis,
       aes(x = wealth_group, y = Stunting, fill = wealth_group)) +
  geom_col() +
  labs(title = "Stunting by Wealth Index",
       y = "Prevalence (%)") +
  theme_minimal()
```


## 6.3 Stunting by Residence
```{r stunting_by_residence}
ggplot(residence_analysis,
       aes(x = residence, y = Stunting, fill = residence)) +
  geom_col() +
  labs(
    title = "Stunting by Residence",
    y = "Prevalence (%)",
    x = ""
  ) +
  theme_minimal()
```


## 6.4 Combined Inequality
```{r combined_inequality}
ggplot(combined_wr,
       aes(x = wealth_group, y = Stunting, fill = residence)) +
  geom_col(position = "dodge") +
  labs(title = "Stunting by Wealth and Residence",
       y = "Prevalence (%)") +
  theme_minimal()
```


## Wasting
```{r wasting}
ggplot(wealth_analysis,
       aes(x = wealth_group, y = Wasting, fill = wealth_group))+
  geom_col()+
  labs(title="Wasting by Wealth", y="Prevalence (%)")+
  theme_minimal()
```


## Underweight
```{r underweight}
ggplot(wealth_analysis,
       aes(x = wealth_group, y = Underweight, fill = wealth_group))+
  geom_col()+
  labs(title="Underweight by Wealth", y="Prevalence (%)")+
  theme_minimal()
```


#     Exporting Project Outputs
  
# prevalence_table.csv
# stunting_plot.png
# wasting_plot.png
# underweight_plot.png

#  1. Create prevalence table
```{r}
prevalence_results <- c(
  Stunting = weighted.mean(DHS_Analysis$stunted,
                           DHS_Analysis$weight,
                           na.rm = TRUE) * 100,
  Wasting = weighted.mean(DHS_Analysis$wasted,
                          DHS_Analysis$weight,
                          na.rm = TRUE) * 100,
  Underweight = weighted.mean(DHS_Analysis$underweight,
                              DHS_Analysis$weight,
                              na.rm = TRUE) * 100
)

prevalence_table <- data.frame(
  Indicator = names(prevalence_results),
  Prevalence = round(as.numeric(prevalence_results), 1)
)

print(prevalence_table)
```

# Save summary table
```{r}
write.csv(prevalence_table,
          "outputs/prevalence_table.csv",
          row.names = FALSE)
```

# 2.Create individual plots

```{r}
library(ggplot2)
```

# (a) Stunting plot
```{r}
stunting_plot <- ggplot(
  subset(prevalence_table, Indicator == "Stunting"),
  aes(x = Indicator, y = Prevalence)
) +
  geom_col(fill = "steelblue", width=0.3) +
  labs(
    title = "Prevalence of Stunting in Kenya",
    y = "Percentage (%)",
    x = ""
  ) +
  theme_minimal()
```

# (b) Wasting plot
```{r}
wasting_plot <- ggplot(
  subset(prevalence_table, Indicator == "Wasting"),
  aes(x = Indicator, y = Prevalence)
) +
  geom_col(fill = "tomato", width=0.3) +
  labs(
    title = "Prevalence of Wasting in Kenya",
    y = "Percentage (%)",
    x = ""
  ) +
  theme_minimal()
```

# (c) Underweight plot
```{r}
underweight_plot <- ggplot(
  subset(prevalence_table, Indicator == "Underweight"),
  aes(x = Indicator, y = Prevalence)
) +
  geom_col(fill = "darkgreen", width=0.3) +
  labs(
    title = "Prevalence of Underweight in Kenya",
    y = "Percentage (%)",
    x = ""
  ) +
  theme_minimal()
```

# One Combined Plot
```{r}
overall_prevalence_plot <- ggplot(
  prevalence_table,
  aes(x = Indicator, y = Prevalence, fill = Indicator)
) +
  geom_col() +
  labs(
    title = "Prevalence of Child Malnutrition",
    y = "Prevalence (%)",
    x = ""
  ) +
  theme_minimal()
```


# Display plots
```{r}
print(stunting_plot)
print(wasting_plot)
print(underweight_plot)
print(overall_prevalence_plot)
```

# Save plots

# Stunting
```{r}
ggsave("outputs/stunting_plot.png",
       plot = stunting_plot,
       width = 8,
       height = 5,
       dpi = 300)
```

# Wasting
```{r}
ggsave("outputs/wasting_plot.png",
       plot = wasting_plot,
       width = 8,
       height = 5,
       dpi = 300)
```

# Underweight
```{r}
ggsave("outputs/underweight_plot.png",
       plot = underweight_plot,
       width = 8,
       height = 5,
       dpi = 300)
```

# Overall_prevalence
```{r}
ggsave("outputs/overall_prevalence_plot.png",
       plot = overall_prevalence_plot,
       width = 8,
       height = 5,
       dpi = 300)
```


#  Conclusion

This analysis estimated the prevalence of stunting, wasting, and underweight
among children in Kenya using DHS data. Results revealed substantial
inequalities by wealth status, place of residence, maternal education,
and sex of the child.

Children from poorer households and rural areas experienced higher levels
of malnutrition, highlighting the importance of targeted nutrition and
public health interventions.
